---
title: "MOURA_LMF2021"
author: "Cândido-Sobrinho, S.A."
date: "24/09/2021"
output: html_document
---

This work is licensed under a [Creative Commons Attribution 4.0 International License][cc-by].

It is made public available at: https://github.com/candidosobrinhosa/MOURA_LMF_2022

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1 Loading libraries
```{r}
base::library(tidyverse)
base::library(FactoMineR)
base::library(factoextra)
base::library(ggpubr)
base::library(ape)
base::library(svglite)
base::library(dendextend)
```


## 2 Initializing lists
```{r}
base::list() -> datasets
base::list() -> pcas
base::list() -> dendrograms
```


## 3 Loading data
```{r}
utils::read.csv("./data/Crambe_Revision1.csv", sep = ";", dec = ".") -> datasets$crambe
utils::read.csv("./data/Raph_Revision1.csv", sep = ";", dec = ".")   -> datasets$raphan
```

### 3.1 Scaling data
```{r}
base::cbind(datasets$crambe[1:3],scale(datasets$crambe[4:46])) -> datasets$crambe_scaled
base::cbind(datasets$raphan[1:3],scale(datasets$raphan[4:46])) -> datasets$raphan_scaled
```

### 3.2 Subsetting data
```{r}
base::list() -> subsets

datasets$crambe_scaled |> filter(DAY == 7)  |> select(!c(DAY)) -> subsets$crambe_scaled_7
datasets$crambe_scaled |> filter(DAY == 14) |> select(!c(DAY)) -> subsets$crambe_scaled_14
datasets$crambe_scaled |> filter(DAY == 21) |> select(!c(DAY)) -> subsets$crambe_scaled_21

datasets$raphan_scaled |> filter(DAY == 7)  |> select(!c(DAY)) -> subsets$raphan_scaled_7
datasets$raphan_scaled |> filter(DAY == 14) |> select(!c(DAY)) -> subsets$raphan_scaled_14
datasets$raphan_scaled |> filter(DAY == 21) |> select(!c(DAY)) -> subsets$raphan_scaled_21
```


## 4 PCAs

### 4.1 Assessing number of PCs
```{r}
FactoMineR::PCA(datasets$crambe_scaled,
    scale = FALSE,
    ncp = 10,
    quali.sup = 1:3,
    graph = F) -> pcas$crambe

FactoMineR::PCA(datasets$raphan_scaled,
    scale = FALSE,
    ncp = 10,
    quali.sup = 1:3,
    graph = F) -> pcas$raphan
```

#### Extracting eigenvalues dataframes.
```{r}
base::as.data.frame(base::cbind(pcas$crambe[["eig"]], c(1:43))) -> pcas$crambe_eig
base::as.data.frame(base::cbind(pcas$crambe[["eig"]], c(1:43))) -> pcas$raphan_eig
```

#### Plotting

#### Crambe
```{r}
pcas$crambe_eig |>

  ggpubr::ggline(x="V4", y="eigenvalue",
         xlab="Number of Principal Components (PCs)",
         ylab="Eigenvalues", color = "black") +

  ggpubr::rotate_x_text()

  ggplot2::ggsave("./results/FigS1_crambe_eigenvalues.svg", device="svg", scale=2, width = 3, height = 2)
```


#### Raphanus
```{r}
pcas$raphan_eig |>

  ggpubr::ggline(x="V4", y="eigenvalue",
         xlab="Number of Principal Components (PCs)",
         ylab="Eigenvalues", color = "black") +

  ggpubr::rotate_x_text()

  ggplot2::ggsave("./results/FigS2_raphan_eigenvalues.svg", device="svg", scale=2, width = 3, height = 2)
```

### 4.2 PCAs
```{r}
FactoMineR::PCA(subsets$crambe_scaled_7,
    scale = FALSE,
    ncp = 4,
    quali.sup = 1:2,
    graph = F) -> pcas$crambe7

FactoMineR::PCA(subsets$crambe_scaled_14,
    scale = FALSE,
    ncp = 4,
    quali.sup = 1:2,
    graph = F) -> pcas$crambe14

FactoMineR::PCA(subsets$crambe_scaled_21,
    scale = FALSE,
    ncp = 4,
    quali.sup = 1:2,
    graph = F) -> pcas$crambe21

FactoMineR::PCA(subsets$raphan_scaled_7,
    scale = FALSE,
    ncp = 4,
    quali.sup = 1:2,
    graph = F) -> pcas$raphan7

FactoMineR::PCA(subsets$raphan_scaled_14,
    scale = FALSE,
    ncp = 4,
    quali.sup = 1:2,
    graph = F) -> pcas$raphan14

FactoMineR::PCA(subsets$raphan_scaled_21,
    scale = FALSE,
    ncp = 4,
    quali.sup = 1:2,
    graph = F) -> pcas$raphan21
```


#### PCAs Biplot (per day); Crambe.
```{r}
factoextra::fviz_pca_biplot(pcas$crambe7, 
                geom.ind = "point",
                fill.ind = subsets$crambe_scaled_7$TREATMENT,
                col.ind = "black",
                pointshape = 21, pointsize = 4,
                addEllipses = TRUE,
                repel = FALSE,
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot",
                legend.title = list(fill = "Treatments", color = "Contribution")
                )

ggplot2::ggsave("./Results/Fig1_crambe7.svg", 
       device="svg",
       scale=2,
       width = 3,
       height = 2
       )

factoextra::fviz_pca_biplot(pcas$crambe14, 
                geom.ind = "point",
                fill.ind = subsets$crambe_scaled_14$TREATMENT,
                col.ind = "black",
                pointshape = 21, pointsize = 4,
                addEllipses = TRUE,
                repel = FALSE,
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot",
                legend.title = list(fill = "Treatments", color = "Contribution")
                )

ggplot2::ggsave("./Results/Fig_crambe14.svg", 
       device="svg",
       scale=2,
       width = 3,
       height = 2
       )

factoextra::fviz_pca_biplot(pcas$crambe21, 
                geom.ind = "point",
                fill.ind = subsets$crambe_scaled_21$TREATMENT,
                col.ind = "black",
                pointshape = 21, pointsize = 4,
                addEllipses = TRUE,
                repel = FALSE,
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot",
                legend.title = list(fill = "Treatments", color = "Contribution")
                )

ggplot2::ggsave("./Results/Fig_crambe21.svg", 
       device="svg",
       scale=2,
       width = 3,
       height = 2
       )
```

#### 4.4 Raphanus
```{r}
factoextra::fviz_pca_biplot(pcas$raphan7, 
                geom.ind = "point",
                fill.ind = subsets$raphan_scaled_7$TREATMENT,
                col.ind = "black",
                pointshape = 21, pointsize = 4,
                addEllipses = TRUE,
                repel = FALSE,
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot",
                legend.title = list(fill = "Treatments", color = "Contribution")
                )

ggplot2::ggsave("./Results/Fig_raphan7.svg", 
       device="svg",
       scale=2,
       width = 3,
       height = 2
       )

factoextra::fviz_pca_biplot(pcas$raphan14, 
                geom.ind = "point",
                fill.ind = subsets$raphan_scaled_14$TREATMENT,
                col.ind = "black",
                pointshape = 21, pointsize = 4,
                addEllipses = TRUE,
                repel = FALSE,
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot",
                legend.title = list(fill = "Treatments", color = "Contribution")
                )

ggplot2::ggsave("./Results/Fig_raphan14.svg", 
       device="svg",
       scale=2,
       width = 3,
       height = 2
       )

factoextra::fviz_pca_biplot(pcas$raphan21, 
                geom.ind = "point",
                fill.ind = subsets$raphan_scaled_21$TREATMENT,
                col.ind = "black",
                pointshape = 21, pointsize = 4,
                addEllipses = TRUE,
                repel = FALSE,
                col.var = "contrib",
                gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                title = "PCA Biplot",
                legend.title = list(fill = "Treatments", color = "Contribution")
                )

ggplot2::ggsave("./Results/Fig_raphan21.svg", 
       device="svg",
       scale=2,
       width = 3,
       height = 2
       )
```

#### 4.5 Variable Contributions

#### Crambe
```{r}
factoextra::fviz_contrib(pcas$crambe7,
             choice = "var",
             axes = 1:2,
             top = 20,
             title="Contribution of variables to PC1 and PC2")

ggplot2::ggsave("./Results/Fig_crambe7_contrib.svg",
       device="svg"
       )

factoextra::fviz_contrib(pcas$crambe14,
             choice = "var",
             axes = 1:2,
             top = 20,
             title="Contribution of variables to PC1 and PC2")

ggplot2::ggsave("./Results/Fig_crambe14_contrib.svg",
       device="svg"
       )

factoextra::fviz_contrib(pcas$crambe21,
             choice = "var",
             axes = 1:2,
             top = 20,
             title="Contribution of variables to PC1 and PC2")

ggplot2::ggsave("./Results/Fig_crambe21_contrib.svg",
       device="svg"
       )
```

#### Raphanus
```{r}
factoextra::fviz_contrib(pcas$raphan7,
             choice = "var",
             axes = 1:2,
             top = 20,
             title="Contribution of variables to PC1 and PC2")

ggplot2::ggsave("./Results/Fig_raphan7_contrib.svg",
       device="svg"
       )

factoextra::fviz_contrib(pcas$raphan14,
             choice = "var",
             axes = 1:2,
             top = 20,
             title="Contribution of variables to PC1 and PC2")

ggplot2::ggsave("./Results/Fig_raphan14_contrib.svg",
       device="svg"
       )

factoextra::fviz_contrib(pcas$raphan21,
             choice = "var",
             axes = 1:2,
             top = 20,
             title="Contribution of variables to PC1 and PC2")

ggplot2::ggsave("./Results/Fig_raphan21_contrib.svg",
       device="svg"
       )
```


### 5 Dendrograms

#### 5.1 Generating matrices
```{r}
base::as.matrix(subsets$crambe_scaled_7[,-(1:2)] ) -> dendrograms$crambe7
base::as.matrix(subsets$crambe_scaled_14[,-(1:2)] ) -> dendrograms$crambe14
base::as.matrix(subsets$crambe_scaled_21[,-(1:2)] ) -> dendrograms$crambe21

base::as.matrix(subsets$raphan_scaled_7[,-(1:2)] ) -> dendrograms$raphan7
base::as.matrix(subsets$raphan_scaled_14[,-(1:2)] ) -> dendrograms$raphan14
base::as.matrix(subsets$raphan_scaled_21[,-(1:2)] ) -> dendrograms$raphan21
```


#### 5.2 Calculating distances
```{r}
stats::hclust(stats::dist(dendrograms$crambe7, method = "euclidean"), method = "ward.D2")  -> dendrograms$crambe7_hc
stats::hclust(stats::dist(dendrograms$crambe14, method = "euclidean"), method = "ward.D2") -> dendrograms$crambe14_hc
stats::hclust(stats::dist(dendrograms$crambe21, method = "euclidean"), method = "ward.D2") -> dendrograms$crambe21_hc

stats::hclust(stats::dist(dendrograms$crambe7, method = "euclidean"), method = "ward.D2")  -> dendrograms$raphan7_hc
stats::hclust(stats::dist(dendrograms$crambe14, method = "euclidean"), method = "ward.D2") -> dendrograms$raphan14_hc
stats::hclust(stats::dist(dendrograms$crambe21, method = "euclidean"), method = "ward.D2") -> dendrograms$raphan21_hc
```

#### Injecting labels into dendrograms
```{r}
subsets$crambe_scaled_7$REPLICATE  -> dendrograms$crambe7_hc$labels
subsets$crambe_scaled_14$REPLICATE -> dendrograms$crambe14_hc$labels
subsets$crambe_scaled_21$REPLICATE -> dendrograms$crambe21_hc$labels

subsets$raphan_scaled_7$REPLICATE  -> dendrograms$raphan7_hc$labels
subsets$raphan_scaled_14$REPLICATE -> dendrograms$raphan14_hc$labels
subsets$raphan_scaled_21$REPLICATE -> dendrograms$raphan21_hc$labels
```

#### 5.3 Plotting (as phylo, important to check clustering)
```{r}
svglite::svglite(filename = "./results/HC_crambe7.svg")
graphics::plot(ape::as.phylo(dendrograms$crambe7_hc), type = "unrooted", no.margin = T)
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_crambe14.svg")
graphics::plot(as.phylo(dendrograms$crambe14_hc), type = "unrooted", no.margin = T)
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_crambe21.svg")
graphics::plot(as.phylo(dendrograms$crambe21_hc), type = "unrooted", no.margin = T)
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_raphan7.svg")
graphics::plot(as.phylo(dendrograms$raphan7_hc), type = "unrooted", no.margin = T)
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_raphan14.svg")
graphics::plot(as.phylo(dendrograms$raphan14_hc), type = "unrooted", no.margin = T)
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_raphan21.svg")
graphics::plot(as.phylo(dendrograms$raphan21_hc), type = "unrooted", no.margin = T)
grDevices::dev.off()
```

#### 5.4 Classical visualization
```{r}
svglite::svglite(filename = "./results/HC_crambe7_2.svg")
dendrograms$crambe7_hc |>
  stats::as.dendrogram() |>
  dendextend::set("branches_k_color", k = 3) |>
  graphics::plot()
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_crambe14_2.svg")
dendrograms$crambe14_hc |>
  stats::as.dendrogram() |>
  dendextend::set("branches_k_color", k = 3) |>
  graphics::plot()
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_crambe21_2.svg")
dendrograms$crambe21_hc |>
  stats::as.dendrogram() |>
  dendextend::set("branches_k_color", k = 3) |>
  graphics::plot()
grDevices::dev.off()


svglite::svglite(filename = "./results/HC_raphan7_2.svg")
dendrograms$raphan7_hc |>
  stats::as.dendrogram() |>
  dendextend::set("branches_k_color", k = 3) |>
  graphics::plot()
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_raphan14_2.svg")
dendrograms$raphan14_hc |>
  stats::as.dendrogram() |>
  dendextend::set("branches_k_color", k = 3) |>
  graphics::plot()
grDevices::dev.off()

svglite::svglite(filename = "./results/HC_raphan21_2.svg")
dendrograms$raphan21_hc |>
  stats::as.dendrogram() |>
  dendextend::set("branches_k_color", k = 3) |>
  graphics::plot()
grDevices::dev.off()
```

